---
title: "Crop Type, Proportion, and Intensity Calculation"
output: html_notebook
---

```{r}
install.packages("sf")
```

```{r setup, message=FALSE, warning=FALSE}
# Load libraries
library(tidyverse)
library(sf)
library(terra)
library(dplyr)
library(stringr)
library(exactextractr)

# Input/output folder paths
RAW_DATA_DIR <- "../data/raw/"
INTERIM_DATA_DIR <- "../data/interim/"
PROCESSED_DATA_DIR <- "../data/processed/"
```

------------------------------------------------------------------------

### **Read Fields Data**

```{r}
fields <- read.csv(file.path(INTERIM_DATA_DIR, "2_nonmissing_annual_averages.csv"))
print(dim(fields))
head(fields)
```

```{r}
unique_fields <- fields %>%
  select(field_fvid, lng, lat) %>%
  distinct()

unique_fields_sf <- unique_fields %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326, remove = FALSE)
head(unique_fields_sf)
```

## **Extract Crop Type for Each Field-Year (Majority Pixel Method)**

First, we extract all unique fields and their geographic locations.  
Then, for each year from 2010 to 2024, we identify the crop type for each field using the majority crop method. This approach uses a small buffer around each field's center to account for cases where the exact center may not represent the true field crop (e.g., due to trees, machinery, or small structures within the field).

The result is an interim dataset containing the most common (majority) crop code for each field and year.  
Note: If multiple crop types tie as the majority, additional rows may be created. In that case, we retain only the first occurrence per field-year.

```{r}
cdl_years = 2010:2024
get_majority_crop <- function(fields_sf, buffer_size = 50, cdl_years = 2011:2012) {
  lapply(cdl_years, function(yr) {
    # Load CDL raster for the given year
    r <- str_glue("../data/cdl/CDL_{yr}_55.tif")
    message("Processing Year: ", yr)
    
    # Buffer the points (for ALL fields at once)
    buffered_fields <- fields_sf %>%
      st_transform(5070) %>%
      st_buffer(buffer_size)
    
    # Extract data with pixel-based majority calculation for all fields at once
    exact_extract(
      rast(r), buffered_fields,
      include_cols = "field_fvid",
      coverage_area = TRUE,
      summarize_df = TRUE,
      fun = function(df) {
        df %>%
          group_by(field_fvid, value) %>%
          summarise(
            pixel_count = n(),
            total_area = sum(coverage_area, na.rm = TRUE),
            .groups = "drop"
          ) %>%
          slice_max(pixel_count, n = 1) %>%
          mutate(buffer_radius = buffer_size) %>%
          select(field_fvid, crop_code = value, pixel_count, total_area, buffer_radius)
      }
    ) %>%
      as_tibble() %>%
      mutate(cdl_year = yr)
  }) %>%
    bind_rows()
}

crop_codes <- get_majority_crop(unique_fields_sf, buffer_size = 50, cdl_years = cdl_years)
dim(crop_codes)
```

```{r}
# if there are several majority ones we need to pick up one
field_id <- crop_codes %>%
  group_by(field_fvid, cdl_year) %>%
  slice(1) %>%
  ungroup()
dim(crop_codes_single)

# Save crop_codes_single to interim directory
write.csv(crop_codes_single, file.path(INTERIM_DATA_DIR, "3_R_fields_crop_type.csv"), row.names = FALSE)

```

## Function to PLOT fields with given year

```{r}
plot_field_buffer <- function(fields_sf, field_id, year, buffer_size, raster_dir = "../data/cdl/") {
  require(terra)
  require(sf)
  require(stringr)
  
  # 1. Select and project the field point
  site <- fields_sf %>%
    filter(field_fvid == field_id) %>%
    st_transform(5070)
  
  # 2. Create buffer (in meters, EPSG:5070)
  buffer <- st_buffer(site, buffer_size)
  
  # 3. Load the CDL raster for the year
  raster_path <- str_glue("{raster_dir}CDL_{year}_55.tif")
  cdl_raster <- rast(raster_path)
  
  # 4. Transform buffer/site to raster CRS for cropping/plotting
  buffer_raster_crs <- st_transform(buffer, crs(cdl_raster))
  site_raster_crs <- st_transform(site, crs(cdl_raster))
  
  # 5. Crop raster to the buffer area
  cdl_crop <- crop(cdl_raster, vect(buffer_raster_crs))
  
  # 6. Plot
  plot(cdl_crop, main = paste("Buffer Zone for Field", field_id, "in Year", year))
  plot(st_geometry(buffer_raster_crs), add = TRUE, border = "blue", lwd = 2)
  plot(st_geometry(site_raster_crs), add = TRUE, col = "red", pch = 20, cex = 2)
}
plot_field_buffer(unique_fields_sf, 62999, 2010, buffer_size = 1000)
```

